Supplementary material

### 2021-12-15 submission/manuscript.rmd
shared version

---
title: |
 Linear interaction of treatment with baseline risk was sufficient for predicting
 individualized benefit
abstract: |
 **Objective**: To compare different risk-based methods predicting
 individualized treatment effects in RCTs. **Study Design and
 Setting**: We simulated RCT data using diverse assumptions for the
 average treatment effect, a baseline prognostic index of risk (PI),
 the shape of its interaction with treatment (none, linear, quadratic
 or non-monotonic) and the magnitude of treatment-related harms. In
 each sample we predicted absolute benefit using: models with a
 constant relative treatment effect; stratification in quarters of the
 PI; models including a linear interaction of treatment with the PI;
 models including an interaction of treatment with a restricted cubic
 spline (RCS) transformation of the PI; an adaptive approach using
 Akaike’s Information Criterion. We evaluated predictive performance
 using root mean squared error and measures of discrimination and
 calibration for benefit. **Results**: The linear-interaction model
 and the RCS-interaction model outperformed the constant treatment
 effect model in many simulation scenarios. The RCS-model was optimal
 when quadratic or non-monotonic deviations from a constant treatment
 effect were stronger, and when sample size was larger. Larger sample
 size also supported the adaptive approach. **Conclusion**: An
 interaction between risk and treatment assignment generally improved
 treatment effect predictions. Non-linear RCS interactions should be
 considered for larger sample size.
author:
  - name: Alexandros Rekkas
    affiliation: 1
  - name: Peter R. Rijnbeek
    affiliation: 1
  - name: Ewout W. Steyerberg
    affiliation: 2
  - name: David van Klaveren
    affiliation: 3
address:
  - code: 1
    address: Department of Medical Informatics, Erasmus Medical Center, Rotterdam, The Netherlands
  - code: 2
    address: Department of Biomedical Data Sciences, Leiden University Medical Center, Leiden, The Netherlands
  - code: 3
    address: Department of Public Health, Erasmus Medical Center, Rotterdam, The Netherlands
numbersections: true
linenumbers: false
journal: arXiv.org
keywords:
  - treatment effect heterogeneity
  - absolute benefit
  - prediction models
output: 
  rticles::elsevier_article:
    keep_tex: true
  bookdown::word_document2:
    reference_docx: reference.docx
geometry: margin=1.0in
date: false
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \usepackage{amsmath}
  - \doublespacing
  - \usepackage{amssymb}
  - \usepackage{bm}
  - \usepackage{caption}
  - \usepackage{booktabs}
  - \date{}
  - \newcommand\given[1][]{\:#1\vert\:}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: jce.csl
---

@@

In comparison with the true approach, dsicrimination for benefit in
the scenario with a constant relative treatment effect was only
slightly lower for the linear interaction model, but more
substantially lower for the non-linear RCS approaches (Figure
\ref{fig:discrimination}; panel A). With strong linear or quadratic
deviations from a constant relative treatment effect, all methods
discriminated quite similarly (Figure \ref{fig:discrimination}; panels
B-C). In the scenario with non-monotonic deviations, the constant
effect model had much lower discriminative ability compared to
non-linear RCS methods (median AUC of
`r aucDistribution %>% filter(scenarioId == 397) %>% mutate(across(-one_of("scenarioId"), ~ d(.x))) %>% pull(median_constant)` 
for the constant effects model,
`r aucDistribution %>% filter(scenarioId == 397) %>% mutate(across(-one_of("scenarioId"), ~ d(.x))) %>% pull(median_linear)`
for the linear interaction model and
`r aucDistribution %>% filter(scenarioId == 397) %>% mutate(across(-one_of("scenarioId"), ~ d(.x))) %>% pull(median_rcs_3)`
for the best-performing RCS-3; Figure \ref{fig:rmseauc}; panel D).
With increasing number of RCS knots, we observed decreasing median
values and increasing variability of the c-for-benefit in all
scenarios. ***With increasing sample size... (Suppl Fig...)***

@@

In terms of calibration for benefit, the constant effects model
outperformed all other models in the scenario with true constant
treatment effects, but was miscalibrated for all deviation scenarios
(Figure \ref{fig:calibration}). The linear interaction model showed
best or close to best calibration across all scenarios and only showed
worse calibration compared to RCS-3 in case of non-monotonic
deviations and treatment-related harms (Figure \ref{fig:calibration};
panel D). The adaptive approach was worse calibrated in scenarios with
strong linear and non-monotonic deviations compared to the linear
interaction model and RCS-3. ***With increasing sample size... (Supp
Fig...)***

>>>>

---
title: |
  An interaction between baseline risk and treatment assignment 
  generally imporoved treatment effect predictions
abstract: |
 **Objective**: To compare different risk-based methods predicting
 individualized treatment effects in RCTs. **Study Design and
 Setting**: We simulated RCT data using diverse assumptions for the
 average treatment effect, a baseline prognostic index of risk (PI),
 the shape of its interaction with treatment (none, linear, quadratic
 or non-monotonic) and the magnitude of treatment-related harms. In
 each sample we predicted absolute benefit using: models with a
 constant relative treatment effect; stratification in quarters of the
 PI; models including a linear interaction of treatment with the PI;
 models including an interaction of treatment with a restricted cubic
 spline (RCS) transformation of the PI; an adaptive approach using
 Akaike’s Information Criterion. We evaluated predictive performance
 using root mean squared error and measures of discrimination and
 calibration for benefit. **Results**: The linear-interaction model
 and the RCS-interaction model outperformed the constant treatment
 effect model in many simulation scenarios. The RCS-model was optimal
 when quadratic or non-monotonic deviations from a constant treatment
 effect were stronger, and when sample size was larger. Larger sample
 size also supported the adaptive approach. **Conclusion**: An
 interaction between risk and treatment assignment generally improved
 treatment effect predictions. Non-linear RCS interactions should be
 considered for larger sample size.
author:
  - name: Alexandros Rekkas
    affiliation: 1
  - name: Peter R. Rijnbeek
    affiliation: 1
  - name: David M. Kent
    affiliation: 2
  - name: Ewout W. Steyerberg
    affiliation: 3
  - name: David van Klaveren
    affiliation: 4
address:
  - code: 1
    address: Department of Medical Informatics, Erasmus Medical Center, Rotterdam, The Netherlands
  - code: 2
    address: Predictive Analytics and Comparative Effectiveness Center, Institute for Clinical Research and Health Policy Studies, Tufts Medical Center, Boston, Massachusetts, USA
  - code: 3
    address: Department of Biomedical Data Sciences, Leiden University Medical Center, Leiden, The Netherlands
  - code: 4
    address: Department of Public Health, Erasmus Medical Center, Rotterdam, The Netherlands
numbersections: true
linenumbers: false
journal: arXiv.org
keywords:
  - treatment effect heterogeneity
  - absolute benefit
  - prediction models
output: 
  rticles::elsevier_article:
    keep_tex: true
  bookdown::word_document2:
    reference_docx: reference.docx
geometry: margin=1.0in
date: false
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \usepackage{amsmath}
  - \doublespacing
  - \usepackage{amssymb}
  - \usepackage{bm}
  - \usepackage{caption}
  - \usepackage{booktabs}
  - \date{}
  - \newcommand\given[1][]{\:#1\vert\:}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: jce.csl
---

@@

In comparison with the true approach, dsicrimination for benefit in
the scenario with a constant relative treatment effect was only
slightly lower for the linear interaction model, but substantially
lower for the non-linear RCS approaches (Figure
\ref{fig:discrimination}; panel A). With strong linear or quadratic
deviations from a constant relative treatment effect, all methods
discriminated quite similarly (Figure \ref{fig:discrimination}; panels
B-C). In the scenario with non-monotonic deviations, the constant
effect model had much lower discriminative ability compared to
all other methods (median AUC of
`r aucDistribution %>% filter(scenarioId == 397) %>% mutate(across(-one_of("scenarioId"), ~ d(.x))) %>% pull(median_constant)` 
for the constant effects model,
`r aucDistribution %>% filter(scenarioId == 397) %>% mutate(across(-one_of("scenarioId"), ~ d(.x))) %>% pull(median_linear)`
for the linear interaction model and
`r aucDistribution %>% filter(scenarioId == 397) %>% mutate(across(-one_of("scenarioId"), ~ d(.x))) %>% pull(median_rcs_3)`
for the best-performing RCS-3; Figure \ref{fig:discrimination}; panel
D). The adaptive approach was unstable in terms of discrimination for
benefit, especially in the presence of treamtnt-related harms. With
increasing number of RCS knots, we observed decreasing median values
and increasing variability of the c-for-benefit in all scenarios. When
we increased the sample size to 17,000 we observed similar trends,
however the performance of all methods was more stable (Supplement,
Figure S4). Finally, when we increased the true prediction AUC to 0.85
the adaptive approach in the case of non-monotonic deviations was,
again, quite unstable, especially with null or moderate
treatment-related harms (Supplement, Figure S5). In these scenarios,
the adaptive approach tended to select more often the much inferior
constant treatment effect method (Supplement, Figure S3)

@@

In terms of calibration for benefit, the constant effects model
outperformed all other models in the scenario with true constant
treatment effects, but was miscalibrated for all deviation scenarios
(Figure \ref{fig:calibration}). The linear interaction model showed
best or close to best calibration across all scenarios and only showed
worse calibration compared to RCS-3 in case of non-monotonic
deviations and treatment-related harms (Figure \ref{fig:calibration};
panel D). The adaptive approach was worse calibrated in scenarios with
strong linear and non-monotonic deviations compared to the linear
interaction model and RCS-3. When we increased sample size to 17,000
similar conclusions on calibration for benefit could be drawn. As
expected, all methods displayed more stable calibration performance
due to the larger number of patients (Supplement, Figure S6). When we
increased the true prediction AUC to 0.85, the linear interaction
model was worse calibrated, on average, than RCS-3 in the case of
strong quadratic deviations from constant relative treatment effects
(Supplement, Figure S7).

### 2021-12-15 submission/manuscript.tex
shared version

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage[margin=1.0in]{geometry}
\bibliographystyle{elsarticle-harv}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Linear interaction of treatment with baseline risk was sufficient for predicting individualized benefit},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

  \title{Linear interaction of treatment with baseline risk was
sufficient for predicting individualized benefit}
    \author[1]{Alexandros Rekkas}

@@

    \author[2]{Ewout W. Steyerberg}

    \author[3]{David van Klaveren}

      \address[1]{Department of Medical Informatics, Erasmus Medical
Center, Rotterdam, The Netherlands}
    \address[2]{Department of Biomedical Data Sciences, Leiden
University Medical Center, Leiden, The Netherlands}
    \address[3]{Department of Public Health, Erasmus Medical Center,
Rotterdam, The Netherlands}

@@

In comparison with the true approach, dsicrimination for benefit in the
scenario with a constant relative treatment effect was only slightly
lower for the linear interaction model, but more substantially lower for
the non-linear RCS approaches (Figure \ref{fig:discrimination}; panel
A). With strong linear or quadratic deviations from a constant relative
treatment effect, all methods discriminated quite similarly (Figure
\ref{fig:discrimination}; panels B-C). In the scenario with
non-monotonic deviations, the constant effect model had much lower
discriminative ability compared to non-linear RCS methods (median AUC of
0.4971 for the constant effects model, 0.5285 for the linear interaction
model and 0.5304 for the best-performing RCS-3; Figure
\ref{fig:rmseauc}; panel D). With increasing number of RCS knots, we
observed decreasing median values and increasing variability of the
c-for-benefit in all scenarios. \textbf{\emph{With increasing sample
size\ldots{} (Suppl Fig\ldots)}}

@@

In terms of calibration for benefit, the constant effects model
outperformed all other models in the scenario with true constant
treatment effects, but was miscalibrated for all deviation scenarios
(Figure \ref{fig:calibration}). The linear interaction model showed best
or close to best calibration across all scenarios and only showed worse
calibration compared to RCS-3 in case of non-monotonic deviations and
treatment-related harms (Figure \ref{fig:calibration}; panel D). The
adaptive approach was worse calibrated in scenarios with strong linear
and non-monotonic deviations compared to the linear interaction model
and RCS-3. \textbf{\emph{With increasing sample size\ldots{} (Supp
Fig\ldots)}}

>>>>

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage[margin=1.0in]{geometry}
\bibliographystyle{elsarticle-harv}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={An interaction between baseline risk and treatment assignment generally imporoved treatment effect predictions},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

  \title{An interaction between baseline risk and treatment assignment
generally imporoved treatment effect predictions}
    \author[1]{Alexandros Rekkas}

@@

    \author[2]{David M. Kent}

    \author[3]{Ewout W. Steyerberg}

    \author[4]{David van Klaveren}

      \address[1]{Department of Medical Informatics, Erasmus Medical
Center, Rotterdam, The Netherlands}
    \address[2]{Predictive Analytics and Comparative Effectiveness
Center, Institute for Clinical Research and Health Policy Studies, Tufts
Medical Center, Boston, Massachusetts, USA}
    \address[3]{Department of Biomedical Data Sciences, Leiden
University Medical Center, Leiden, The Netherlands}
    \address[4]{Department of Public Health, Erasmus Medical Center,
Rotterdam, The Netherlands}

@@

In comparison with the true approach, dsicrimination for benefit in the
scenario with a constant relative treatment effect was only slightly
lower for the linear interaction model, but substantially lower for the
non-linear RCS approaches (Figure \ref{fig:discrimination}; panel A).
With strong linear or quadratic deviations from a constant relative
treatment effect, all methods discriminated quite similarly (Figure
\ref{fig:discrimination}; panels B-C). In the scenario with
non-monotonic deviations, the constant effect model had much lower
discriminative ability compared to all other methods (median AUC of
0.4971 for the constant effects model, 0.5285 for the linear interaction
model and 0.5304 for the best-performing RCS-3; Figure
\ref{fig:discrimination}; panel D). The adaptive approach was unstable
in terms of discrimination for benefit, especially in the presence of
treamtnt-related harms. With increasing number of RCS knots, we observed
decreasing median values and increasing variability of the c-for-benefit
in all scenarios. When we increased the sample size to 17,000 we
observed similar trends, however the performance of all methods was more
stable (Supplement, Figure S4). Finally, when we increased the true
prediction AUC to 0.85 the adaptive approach in the case of
non-monotonic deviations was, again, quite unstable, especially with
null or moderate treatment-related harms (Supplement, Figure S5). In
these scenarios, the adaptive approach tended to select more often the
much inferior constant treatment effect method (Supplement, Figure S3)

@@

In terms of calibration for benefit, the constant effects model
outperformed all other models in the scenario with true constant
treatment effects, but was miscalibrated for all deviation scenarios
(Figure \ref{fig:calibration}). The linear interaction model showed best
or close to best calibration across all scenarios and only showed worse
calibration compared to RCS-3 in case of non-monotonic deviations and
treatment-related harms (Figure \ref{fig:calibration}; panel D). The
adaptive approach was worse calibrated in scenarios with strong linear
and non-monotonic deviations compared to the linear interaction model
and RCS-3. When we increased sample size to 17,000 similar conclusions
on calibration for benefit could be drawn. As expected, all methods
displayed more stable calibration performance due to the larger number
of patients (Supplement, Figure S6). When we increased the true
prediction AUC to 0.85, the linear interaction model was worse
calibrated, on average, than RCS-3 in the case of strong quadratic
deviations from constant relative treatment effects (Supplement, Figure
S7).

### 2021-12-15 submission/supplement.rmd
shared version

```{r discriminationSampleSize, cache=FALSE, echo=FALSE, fig.cap="Sample size", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/discrimination_moderate_sample_size.tiff")))
# knitr::include_graphics(here::here("figures/calibration_base.png"))
```

```{r discriminationAuc, cache=FALSE, echo=FALSE, fig.cap="Auc", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/discrimination_moderate_auc.tiff")))
# knitr::include_graphics(here::here("figures/calibration_base.png"))
```

```{r calibrationSampleSize, cache=FALSE, echo=FALSE, fig.cap="Calibration Sample size", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/calibration_moderate_sample_size.tiff")))
# knitr::include_graphics(here::here("figures/calibration_base.png"))
```

```{r calibrationAuc, cache=FALSE, echo=FALSE, fig.cap="Calibration Auc", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/calibration_moderate_auc.tiff")))
# knitr::include_graphics(here::here("figures/calibration_base.png"))
```

>>>>

```{r discriminationSampleSize, cache=FALSE, echo=FALSE, fig.cap="Discrimination for benefit of the considered methods across 500 replications calculated in a simulated sample of size 500,000. True prediction AUC of 0.75 and sample size of 17,000", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/discrimination_moderate_sample_size.tiff")))
# knitr::include_graphics(here::here("figures/calibration_base.png"))
```

```{r discriminationAuc, cache=FALSE, echo=FALSE, fig.cap="Discrimination for benefit of the considered methods across 500 replications calculated in a simulated sample of size 500,000. True prediction AUC of 0.85 and sample size of 4,250", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/discrimination_moderate_auc.tiff")))
# knitr::include_graphics(here::here("figures/calibration_base.png"))
```

```{r calibrationSampleSize, cache=FALSE, echo=FALSE, fig.cap="Calibration for benefit of the considered methods across 500 replications calculated in a simulated sample of size 500,000. True prediction AUC of 0.75 and sample size of 17,000", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/calibration_moderate_sample_size.tiff")))
# knitr::include_graphics(here::here("figures/calibration_base.png"))
```

```{r calibrationAuc, cache=FALSE, echo=FALSE, fig.cap="Calibration for benefit of the considered methods across 500 replications calculated in a simulated sample of size 500,000. True prediction AUC of 0.85 and sample size of 4,250", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/calibration_moderate_auc.tiff")))
# knitr::include_graphics(here::here("figures/calibration_base.png"))
```
